### 2012-10-19 thesis/MA-Thesis.tex
* increased font size, from 10 to 12pt * increased linespacing, now: 1,5 * Chapter "Design" is now splittet into "Visual formalisms" and   "Automated constraint generation" * Section "6.1 Operator constraints" now filled (with lots of pictures) * smaller changes due to new font size and linespacing

\documentclass[10pt,a4paper,headsepline,footsepline,DIV13,BCOR12mm]{scrbook}
\usepackage[english,ngerman]{babel}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{scrpage2}
\pagestyle{scrheadings}

@@

@@

%PREFACE
\selectlanguage{ngerman}
\input{titlepage} 
\newpage
%select english as language!
\selectlanguage{english}
%\pagenumbering{roman}
\pagenumbering{arabic}
\tableofcontents
\newpage
\listoffigures
\newpage
\listoftables
%\pagenumbering{arabic}

>>>>

\documentclass[12pt,a4paper,headsepline,footsepline,DIV13,BCOR12mm]{scrbook}
\usepackage[english,ngerman]{babel}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{scrpage2}
\pagestyle{scrheadings}

@@

\usepackage{setspace}

@@

%PREFACE
\selectlanguage{ngerman}
\input{titlepage}

\onehalfspacing
\newpage
\thispagestyle{empty}

\centerline{\bfseries ERKLÄRUNG}

\vspace{5cm}
Hiermit versichere ich, dass ich diese Masterarbeit selbständig verfasst habe.
Ich habe dazu keine anderen als die angegebenen Quellen und Hilfsmittel
verwendet.

\vspace{1cm}
\begin{flushleft}
%select german for formatting the date
\selectlanguage{ngerman}
Augsburg, den \today \hfill Ludwig Nägele
\end{flushleft}

\newpage
\thispagestyle{empty}
\mbox{}

\newpage

%select english as language!
\selectlanguage{english}
%\pagenumbering{roman}
\pagenumbering{arabic}
\tableofcontents
\newpage
\listoffigures
\newpage
\listoftables
%\pagenumbering{arabic}

### 2012-10-19 thesis/tex/2_introduction.tex
* increased font size, from 10 to 12pt * increased linespacing, now: 1,5 * Chapter "Design" is now splittet into "Visual formalisms" and   "Automated constraint generation" * Section "6.1 Operator constraints" now filled (with lots of pictures) * smaller changes due to new font size and linespacing

Safety critical robot applications require extensive testing or formal verification in order to achieve adequate safe and predictable behavior.
Even if a program does not control actuators itself and thus cannot directly cause any damage, its output might still be a trigger for dangerous actions involving several actors.
Imagine a medication reminder application running on a service robot which guides a person through the process of medication intake, such as the healthcare robotics project at the University of Auckland, New Zealand \cite{jayawardena12:_desig}. All interaction is guided by showing textual instructions on a
\begin{wrapfigure}{r}{0.58\textwidth}
  \centering
  \includegraphics[width=0.56\textwidth]{cafero_being_used}
  \caption{Service robot being used by older person~\cite{robostudio}.}
  \label{fig:cafero_being_used}
\end{wrapfigure}
display, and by speech generation. An error in the program logic could cause an already taken medication to be reminded again and may injure a patient with Alzheimer's disease by causing a medication overdose.
Scientists are working on applications such as the medication reminder~\cite{p11:_feasib} which run on mobile robots serving in nursing homes for the elderly. Equipped with a touch display and several external devices such as a blood pressure measurement tool the robots aim to support and entertain the elderly in their daily activities~\cite{jayawardena12:_desig,5649910}.

@@

Unfortunately the supply of an easy-to-use editor for constraint creation doesn't guarantee adequate thinking about sufficent constraints by the developer. Unmotivated or just unexperienced developers may miss important constraints needed in order to achieve a certain safety level.
To support the user in finding significant constraints it might be useful to automatically compute constraint suggestions which make true testimonies about the current designed program. First of all the generated constraints provide an opportunity for the developer to identify reasonable constraints as well as contradictions between constraints and specification. The latter would mean there are errors in the program which lead to undesired constraints. Furthermore once constraints are created and checked for sanity they can be validated after every program change and thus ensure integrity during the development process. In section~\ref{sec:automatedconstraintgeneration} we introduce a heuristic for finding reasonable safety constraints based on state machine descriptions that describe robots' behaviors.

@@

This work starts with background information about the healthcare domain, short explanations of common safety mechanisms and insights into related work in chapter~\ref{chap:background}. After the goals are mentioned in chapter~\ref{chap:goals}, design decisions are introduced in chapter~\ref{chap:design}.
A test scenario is described in chapter~\ref{chap:testscenarioandevaluation}, where the presented tools get evaluated.
Chapter~\ref{chap:conclusionandfuturework} gives a short conclusion and an outlook how the work presented in this paper could be continued.

>>>>

Safety critical robot applications require extensive testing or formal verification in order to achieve adequate safe and predictable behavior.
Even if a program does not control actuators itself and thus cannot directly cause any damage, its output might still be a trigger for dangerous actions involving several actors.
Imagine a medication reminder application running on a service robot which guides a person through the process of medication intake, such as the healthcare robotics project at the University of Auckland, New Zealand \cite{jayawardena12:_desig}. All interaction is guided by showing textual instructions 
\begin{wrapfigure}{r}{0.58\textwidth}
  \centering
  \includegraphics[width=0.56\textwidth]{cafero_being_used}
  \caption{Service robot being used by older person~\cite{robostudio}.}
  \label{fig:cafero_being_used}
\end{wrapfigure}
on a display, and by speech generation. An error in the program logic could cause an already taken medication to be reminded again and may injure a patient with Alzheimer's disease by causing a medication overdose.
Scientists are working on applications such as the medication reminder~\cite{p11:_feasib} which run on mobile robots serving in nursing homes for the elderly. Equipped with a touch display and several external devices such as a blood pressure measurement tool the robots aim to support and entertain the elderly in their daily activities~\cite{jayawardena12:_desig,5649910}.

@@

Unfortunately the supply of an easy-to-use editor for constraint creation doesn't guarantee adequate thinking about sufficent constraints by the developer. Unmotivated or just unexperienced developers may miss important constraints needed in order to achieve a certain safety level.
To support the user in finding significant constraints it might be useful to automatically compute constraint suggestions which make true testimonies about the current designed program. First of all the generated constraints provide an opportunity for the developer to identify reasonable constraints as well as contradictions between constraints and specification. The latter would mean there are errors in the program which lead to undesired constraints. Furthermore once constraints are created and checked for sanity they can be validated after every program change and thus ensure integrity during the development process. In chapter~\ref{chap:automatedconstraintgeneration} we introduce a heuristic for finding reasonable safety constraints based on state machine definitions that describe robots' behaviors.

@@

This work starts with background information about the healthcare domain, short explanations of common safety mechanisms and insights into related work in chapter~\ref{chap:background}. After the goals are mentioned in chapter~\ref{chap:goals}, the visual formalisms and the automated constraint generation are introduced in chapters~\ref{chap:visualformalisms} and~\ref{chap:automatedconstraintgeneration}.
After a prototype implementation is demonstrated in chapter~\ref{chap:theltlcreatorprototype}, a test scenario is described in chapter~\ref{chap:testscenarioandevaluation} where the presented tools get evaluated.
Chapter~\ref{chap:conclusionandfuturework} gives a short conclusion and an outlook how the work presented in this paper could be continued.

### 2012-10-19 thesis/tex/5_design.tex
* increased font size, from 10 to 12pt * increased linespacing, now: 1,5 * Chapter "Design" is now splittet into "Visual formalisms" and   "Automated constraint generation" * Section "6.1 Operator constraints" now filled (with lots of pictures) * smaller changes due to new font size and linespacing

\chapter{Design}
\label{chap:design}

%On our way towards an useful and simple concept for the defining of safety criteria

%wichtig:
%1. VL
%2. Constraint generierung
%3. integrierung in Robostudio

%different ideas came up

%In this chapter we will intruduce the final design of the visual language and compare it to a first idea of a visual formalism.

In this chapter we introduce a first approach of a visual formalism and later state the final desing of the visual language. Furthermore the idea of automated constraint generation is presented in section~\ref{sec:automatedconstraintgeneration}. The assembly of all the previous features in one tool is described in section~\ref{sec:coalescence}. However, the tool's functionality shall also be available in Robostudio in order to provide safety mechanisms for the healthcare service robots. Section~\ref{sec:integrationintorobostudio} gives a main idea how an integration into Robostudio ought to be.

%The challenge now is to find a similar intuitive and nonmathematical way of representing reasonable safety constraints for the healthbot domain and to figure out how to translate it to corresponding textual expressions. The new concept will be integrated into Robostudio, a design tool for statemachines, in order to enable the user to create safety constraints visually. A common LTL model checker will be added to it as well which fulfils the verification of the translated constraints on the designed statemachine. In case of verification failures the concerning constraints and a visual counter example can be presented.

\section{Visual formalisms}
\label{sec:visualformalisms}

One main difficulty in finding a proper visual formalism was to glean the right level between abstraction and expressiveness since this two characteristics appeared to be mutually exclusive in some points. A first approach, so called template constraints, and its disadvantages are described in section~\ref{sec:templateconstraints}. Subsequently an improved design is presented, which tends to a more abstract direction.

\subsection{Template constraints}
\label{sec:templateconstraints}

What visual presentation of constraints would suit a healthbot application developer and what are the main types of safety constraints he could want to check?
% TODO: wir haben gedacht, dass es einfach zu benutzen ist
Leaded by this question, we developed a visual language based on templates. These templates make different propositions about states and their relations. The user can choose from several graphical constraint types and customize them by parameterizing state variables. So far there are two different kinds of template constraints:

@@

This visual language was presented to potential users for evaluation and feedbak. Soon it turned out that there are some problems and disagreements. The presented constraints were interpreted differently, the semantic seemd to be unclear and not unique.
A try to solve this problem was to equipp both constraint templates with textual explanations which are updated automatically whenever there is a change in parameterization. These additional expressions are also shown in figure~\ref{fig:templateconstraints}.
But unfortunately even the textual eplanations don't seem to be a satisfying solution. For example, one question came up whether the first constraint template allows state (x) to be visited more than once before state (z) is reached.

@@

\subsection{Operator constraints}
\label{sec:operatorconstraints}

@@

\section{Automated constraint generation}
\label{sec:automatedconstraintgeneration}

As already mentioned before, it might be difficult to find all reasonable safety constraints needed for full safety covery. Hence it also can happen, that a prior software version was correct and over time mistakes crept in because of minor adjustments once in a while.
To help prevent this to happen, we introduce the concept of automated constraint generation for state machines. Once a program reaches a complete condition, constraint suggestions which are valid on the current program can be computed fully automatically. On the one hand they may help the developer identifying reasonable constraints he didn't think about as well as contradictions between program and specification, eg. if a computed constraint doesn't make sense at all. On the other hand all constraints can be revalidated after each change to the software and thus ensure consistency.

In order to support the developer in developing safe programs, automated generation and suggestion of constraints might be helpful. But it's helpful only if the suggested constraints are relevant and reasonable. This however depends on the domain and use case. It has to be mentioned that our approach presented in this work has a special focus on the healtcare use cases and may not be applicable for other domains.
Regarding the important constraints postulated in chapter~\ref{chap:goals}, all the relevant states fall into one of the following categories:

@@

	  \node[statein] (1) {$\alpha$};
	  \node[state] (2) [above left of=1] {};
	  \node[state] (3) [below left of=1] {};
	  \node[statein] (4) [above right of=1] {};
	  \node[statein] (5) [below right of=1] {};
	  \node[statein] (8) [right of=5] {$\gamma$};
	  \node[statein] (6) [right of=4] {$\gamma$};
	  \node[statein] (7) [below right of=4] {$\gamma$};

	  \node[statein] (9) [right of=7] {$\beta$};
	  \node[state] (10) [above of=9] {};
	  \node[state] (11) [above right of=9] {};
	  \node[state] (12) [below right of=9] {};

@@

%	  \node[statein] (1) {$\alpha$};
%	  \node[state] (2) [above left of=1] {};
%	  \node[state] (3) [above right of=1] {};
%	  \node[statein] (4) [below left of=1] {};
%	  \node[statein] (6) [below of=4] {$\gamma$};
%	  \node[statein] (7) [below right of=4] {$\gamma$};
%	  \node[statein] (5) [below right of=1] {};
%	  \node[statein] (8) [below of=5] {$\gamma$};
%	  \node[statein] (9) [below of=7] {$\beta$};
%	  \node[state] (10) [right of=9] {};
%	  \node[state] (11) [below left of=9] {};
%	  \node[state] (12) [below right of=9] {};
%	  
%	  \path[dashed]
%	    (2) edge node [left] {} (1)
%	    (3) edge node [left] {} (1)
%	    (10) edge node [left] {} (9)
%	    (9) edge node [left] {} (11)
%	    (9) edge node [left] {} (12);
%		\path[]
%	  	(1) edge node [left] {} (4)
%	        edge node [left] {} (5)
%	    (4) edge node [left] {} (6)
%	        edge node [left] {} (7)
%	    (5) edge node [left] {} (8)
%	    (6) edge node [left] {} (9)
%	    (7) edge node [left] {} (9)
%	    (8) edge node [left] {} (9)
%	    ;

@@

A graphical example of such a subgraph is shown in Fig.~\ref{fig:subgraph}. Using this specification all possible such subgraphs can be detected within a state machine. If the subgraph contains neither inner (infinite) loops nor final states (states having no outgoing transition) we can derive the following valid LTL formulas for each subgraph (with $\alpha$ as proposition for ``$\alpha$ is current state'' and similarly for $\beta$ and $\gamma$):

@@

\section{Coalescence}
\label{sec:coalescence}

Both features the visual language presented in section~\ref{sec:operatorconstraints} and the automated constraint generation in section~\ref{sec:automatedconstraintgeneration} shall be assembled in one tool, the LTLCreator. It shall provide a dashboart with the possibility to manage several constraints, such as a tab panel, and displays for indicating validation results. Besides a button is needed for triggering constraint generation.
The LTLCreator should be available for other state machine based programming environments for integration.
The implementation as a java swing component and a clear API enables the tool for a wide range of use.

Figure~\ref{fig:coalescence} points out how interaction between LTLCreator and programming environments where it is integrated in can be. The programming environment provides the state machine model and notifies the LTLCreator component about all changes made to the program. It will automatically revalidate all existing constraints and provide the up-to-date model for constraint generation.
For load and save functionality, all existing constraints can be received and new ones can be added to the LTLCreator component programmatically.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.7\linewidth]{coalescence}
  \caption{Interaction between LTLCreator and programming environment.}
  \label{fig:coalescence}
\end{figure}

TODO: NuSMV hier auch wichtig?

\section{Integration into Robostudio}
\label{sec:integrationintorobostudio}

Robostudio uses NetBeans as a platform, and all tools and features are organized in windows. Thus also the LTLCreator has to be put into a window and added to the Robostudio platform. Since the defining of safety constraints is an important part while developing service robot behavior, the LTLCreator window shall appear in the center of the platform.

>>>>

%\chapter{Design}
%\label{chap:design}
%
%
%In this chapter we introduce a first approach of a visual formalism and later state the final desing of the visual language. Furthermore the idea of automated constraint generation is presented in section~\ref{sec:automatedconstraintgeneration}. The assembly of all the previous features in one tool is described in section~\ref{sec:coalescence}. However, the tool's functionality shall also be available in Robostudio in order to provide safety mechanisms for the healthcare service robots. Section~\ref{sec:integrationintorobostudio} gives a main idea how an integration into Robostudio ought to be. 
%
%%The challenge now is to find a similar intuitive and nonmathematical way of representing reasonable safety constraints for the healthbot domain and to figure out how to translate it to corresponding textual expressions. The new concept will be integrated into Robostudio, a design tool for statemachines, in order to enable the user to create safety constraints visually. A common LTL model checker will be added to it as well which fulfils the verification of the translated constraints on the designed statemachine. In case of verification failures the concerning constraints and a visual counter example can be presented.

\chapter{Visual formalisms}
\label{chap:visualformalisms}

In this chapter we introduce two different approaches of visual formalisms and later state the final desing of the visual language.
One main difficulty in finding a proper visual formalism is to glean the right level between abstraction and expressiveness since this two characteristics appear to be mutually exclusive in some points. A first approach, so called template constraints and its pros and cons are described in section~\ref{sec:templateconstraints}. Subsequently an improved design is presented, which tends to a more abstract direction.

\section{Template constraints}
\label{sec:templateconstraints}

%It is not a trivial issue to find a
What visual presentation of constraints would suit a healthbot application developer and what are the main types of safety constraints he could want to check?
% TODO: wir haben gedacht, dass es einfach zu benutzen ist
Leaded by this question, we developed a visual language based on templates. These templates make different propositions about states and their relations. The user can choose from several graphical constraint types and customize them by parameterizing state variables. So far there are two different kinds of template constraints:

@@

This visual language was presented to potential users for evaluation and feedbak. Though most of them named it a good concept, soon it turned out that there are some problems and disagreements. The presented constraints were interpreted differently, the semantic seemd to be unclear and not unique.
A try to solve this problem was to equip both constraint templates with textual explanations which are updated automatically whenever there is a change in parameterization. These additional expressions are also shown in figure~\ref{fig:templateconstraints}.
But unfortunately even the textual eplanations don't seem to be a satisfying solution. For example, one question came up whether the first constraint template allows state (x) to be visited more than once before state (z) is reached.

@@

\section{Operator constraints}
\label{sec:operatorconstraints}

@@

%\section{Automated constraint generation}
%\label{sec:automatedconstraintgeneration}
\chapter{Automated constraint generation}
\label{chap:automatedconstraintgeneration}

As already mentioned before, it might be difficult for developers to find all reasonable safety constraints by hand which are needed for full safety covery. Hence it also can happen, that a prior software version was correct and over time mistakes crept in because of minor adjustments once in a while.
To help prevent this to happen, we introduce the concept of automated constraint generation for state machines. Once a program reaches a complete condition, constraint suggestions which are valid on the current program can be computed fully automatically. On the one hand they may help the developer identifying reasonable constraints he didn't think about as well as contradictions between program and specification, eg. if a computed constraint doesn't make sense at all. On the other hand all constraints can be revalidated after each change to the software and thus ensure consistency.

In order to support the developer in developing safe programs, automated generation and suggestion of constraints might be helpful. But it's helpful only if the suggested constraints are relevant and reasonable.

\section{Sub graph definition}

Whether suggested constraints are relevant and reasonable
%This however
depends on the domain and use case. It has to be mentioned that our approach of finding an algorithm for constraint finding presented in this work has a special focus on the healtcare use cases and may not be applicable for other domains.
Regarding the important constraints postulated in chapter~\ref{chap:goals}, we found out that all all the relevant states fall into one of the following categories:

@@

%	  \node[statein] (1) {$\alpha$};
%	  \node[state] (2) [above left of=1] {};
%	  \node[state] (3) [below left of=1] {};
%	  \node[statein] (4) [above right of=1] {};
%	  \node[statein] (5) [below right of=1] {};
%	  \node[statein] (8) [right of=5] {$\gamma$};
%	  \node[statein] (6) [right of=4] {$\gamma$};
%	  \node[statein] (7) [below right of=4] {$\gamma$};
%	  
%	  \node[statein] (9) [right of=7] {$\beta$};
%	  \node[state] (10) [above of=9] {};
%	  \node[state] (11) [above right of=9] {};
%	  \node[state] (12) [below right of=9] {};
%	  
%	  \path[dashed]
%	    (2) edge node [left] {} (1)
%	    (3) edge node [left] {} (1)
%	    (10) edge node [left] {} (9)
%	    (9) edge node [left] {} (11)
%	    (9) edge node [left] {} (12);
%		\path[]
%	  	(1) edge node [left] {} (4)
%	        edge node [left] {} (5)
%	    (4) edge node [left] {} (6)
%	        edge node [left] {} (7)
%	    (5) edge node [left] {} (8)
%	    (6) edge node [left] {} (9)
%	    (7) edge node [left] {} (9)
%	    (8) edge node [left] {} (9)
%	    ;

	  \node[statein] (1) {$\alpha$};
	  \node[state] (2) [above left of=1] {};
	  \node[state] (3) [above right of=1] {};
	  \node[statein] (4) [below left of=1] {};
	  \node[statein] (6) [below of=4] {$\gamma$};
	  \node[statein] (7) [below right of=4] {$\gamma$};
	  \node[statein] (5) [below right of=1] {};
	  \node[statein] (8) [below of=5] {$\gamma$};
	  \node[statein] (9) [below of=7] {$\beta$};
	  \node[state] (10) [right of=9] {};
	  \node[state] (11) [below left of=9] {};
	  \node[state] (12) [below right of=9] {};

@@

@@

A graphical example of such a subgraph is shown in Fig.~\ref{fig:subgraph}. Using this specification all possible such subgraphs can be detected within a state machine.

\section{Constraint determination}

If the subgraph contains neither inner (infinite) loops nor final states (states having no outgoing transition) we can derive the following valid LTL formulas for each subgraph (with $\alpha$ as proposition for ``$\alpha$ is current state'' and similarly for $\beta$ and $\gamma$):

@@

### 2012-10-19 thesis/tex/6_implementation.tex
* increased font size, from 10 to 12pt * increased linespacing, now: 1,5 * Chapter "Design" is now splittet into "Visual formalisms" and   "Automated constraint generation" * Section "6.1 Operator constraints" now filled (with lots of pictures) * smaller changes due to new font size and linespacing

\chapter{The LTLCreator prototype}

As stated in chapter~\ref{chap:goals}, appropriate support by a graphical editor is important for the usefulness of the developed visual language. We therefore designed and implemented a prototype of such an editor. The proposed constraint generation heuristics is integrated in this prototype as well and could thus be evaluated. The implementation is a java swing component, making it easy to integrate the visual language into other java applications. An API enables an underlying state machine which also influences the state propositions available for constraint editing. Furthermore a particular model checker can be specified for constraint validation. As a default, the symbolic model checker NuSMV~\cite{springerlink:10.1007/s100090050046,NuSMV2} is used.

Fig.~\ref{fig:editor} shows a snapshot of the editor's environment: All functionality needed for constraint editing is provided in the tool bar on the right. It contains draggable elements for creating all operator and proposition types as well as a trashcan for deleting. Constraints can be composed in the dashboard in the center of the editor by drag\&drop. The tab functionality on the left allows multiple constraints to be managed. Each tab shows a small thumbnail of the constraint and a symbol indicating its validity (a warning shield for ``incomplete'', a green shield for ``valid'', a red cross for ``invalid'' or an animated ring for ``validation in progress.'') The magic wand button within the tab pane triggers the constraint generation.

TODO: Kleine Inhaltsangabe

\section{Operator constraints}
\section{Automated constraint generation}

@@

\section{Model checker}

@@

\section{Interaction with Robostudio}

>>>>

\chapter{The LTLCreator prototype}
\label{chap:theltlcreatorprototype}

As stated in chapter~\ref{chap:goals}, appropriate support by a graphical editor is important for the usefulness of the developed visual language. We therefore designed and implemented a prototype of such an editor which contains the visual language as well as the proposed constraint generation heuristics. The implementation is a java swing component and has a simple API definition, making it easy to integrate the LTLCreator tool into other java applications.

First of all some details about the implementation of the visual language are given in section~\ref{sec:prototype:operatorconstraints}, directly followed by an explanation about the realization of the automated constraint generation functionality in section~\ref{sec:prototype:automatedconstraintgeneration}.
The assembly of all the previous features in one tool is described in section~\ref{sec:coalescence}. However, the tool's functionality shall also be available in Robostudio in order to provide safety mechanisms for the healthcare service robots and to be a prototype for evaluation. Section~\ref{sec:integrationintorobostudio} gives a main idea how an integration into Robostudio ought to be.

\section{Operator constraints}
\label{sec:prototype:operatorconstraints}

Each of the different operator types, i.e. IfThenOperator, AndOperator, OrOperator, NotOperator, NextOperator, FutureOperator, AlwaysOperator, UntilOperator and StateOperator, has got its own representing class. All of them inherit from one abstract class AbstractOperator which provides the basic functionality beeing used by all operators. The AbstractOperator class is shown in figure~\ref{fig:abstractoperator} and its methods are described as follows:

\begin{description}
	\item[getLTL():] This function returns the LTL formula of an operator and all its children recursively. Applied to the most outer operator it returns the complete formula for the constraint. This function is abstract and implemented by each operator type separately.
	\item[getColor():] Each operator type has got is own significant color which can be retrieved over the abstract function getColor().
	\item[createNewInstance():] In order to provide operator creation over a tool bar, this function instantiates a new operator object dependet on the implementation.
	\item[setMouseOver(boolean):] Mouse over actions cause the underlying operator to be highlighted. This function makes all operators except the hovered one appearing bleeched out.
	\item[paintOperator(Graphics, ...):] Each operator class has to implement the abstract function paintOperator(). It manages all color and shape painting for the respective operator. Additional parameters such as recursive painting including children can be spezified.
	\item[contains(int, int):] Given two coordinates x and y, it can be determined wheter the herewith defined point lays within an operator.
	\item[isSimilar(AbstractOperator):] Two operators can be compared for semantic equality. That means same operators with same children. This functionality is important for avoiding duplicate constraints during constraint generation.
	\item[addChangeListener(OperatorChangeListener):] Whenever there is a change to the operator or one of its children, an operator change listener gets notified about it. The hierarchical architecture of AbstractOperators propagates all change events to the root operator. The function addChangeListener() allows registering for such change notifications.
	\item[removeChangeListener(OperatorChangeListener):] According to addChangeListener this function allows deregistration again.
\end{description}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.6\linewidth]{AbstractOperator} 
  \caption{AbstractOperator class definition.}
  \label{fig:abstractoperator}
\end{figure}

Whereas StateOperators have got no operator children and are leafs in the syntax tree of LTL formulas, all other operator types are either unary or binary operators with respective one or two possible children. 
The graphical representation of both unary and binary operators thus needs to have place holders and docking stations for sub operators. These are realized by buckets which display a small ``drop here'' label and allow operator adding or removing by mouse drops. For this a class Bucket is defined with get and set methods for defining a sub operator.
Figure~\ref{fig:klassendiagramm_operators} depicts the relationship between AbstractOperator, Bucket and operator implementations.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{klassendiagramm_operators} 
  \caption{Class diagram of operator structure.}
  \label{fig:klassendiagramm_operators}
\end{figure}

Once operators are composed in a way that no empty buckets are left over, we retrieve a complete constraint such as the one shown in figure~\ref{fig:operator_tree}. This diagram illustrates the correlation of nested operators and buckets.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{operator_tree} 
  \caption{Instantiation of a constraint with several operators.}
  \label{fig:operator_tree}
\end{figure}

All graphical components such as buckets and operators are implemented as java swing JComponent. Thereby a lot of functionality for the visual configuration is already available such as the complex layout and repaint concept and the container functionality which is used for unary and binary operators as well as buckets.

%PAINT

For lightweight components swing allows the implementation of a paint() method which is responsible for rendering the component. This is used for giving every operator its special look. We hereby made use of swing's Graphics2D functionality what gives a great support in painting rounded shapes and color gradients. An operator is displayed as a rounded rectangle with dents around sub operators and has got a flowing backgound, borders and labels in an operator type specific color. For the significant shape multiple rounded rectangles with and without border are arranged in a way that the result appears to be one peace. Figure~\ref{fig:paintprocess} demonstrates how the process of painting is: For painting the if operator first of all a blank rectangle with border is drawn, directly followed by two filled rectangles with border which are placed where later the two child operators shall be. In the third step the same rectangle as from step one just this time without border is painted on top. This makes all crossing borders disappear. Only labels and buckets have to be added yet and a complete IfThenOperator emerges.

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{paintprocess} 
  \caption{Steps of painting an operator.}
  \label{fig:paintprocess}
\end{figure}

%Um die paint-Logik muss sich nicht gekümmert werden, lediglich die paint mehtode muss implementiert werden, die dem jeweiligen Operator sein spezielles Aussehen verleiht. Dabei wurde auf die Funktionen des Graphics2D zurückgegriffen, was die Erstellung von runden Formen (shapes) und Farbverläufen (stroke?) ermöglicht. Jeder Operatortyp hat eine spezifische Farbe, die für den Farbverlauf im Hintergrund, für den Border und als Textfarbe verwendet wird. Die shapes der Operatoren wurden so umgesetzt, dass sie sich um die enthaltenen Operatoren schlingen (schmiegen). Hierfür wurden auschließlich leere und gefüllte abgerundete Rechtecke gezeichnet, die in der richtigen Reihenfolge übereinandergelegt die gewünschte Endform für den Operator ergibt. grafisches beispiel?

%LAYOUT

Swing also provides a mechanism for layouting components, that means depending on child components the aspired sizes and positions of components can be retrieved. As known from figure~\ref{fig:operators} in section~\ref{sec:operatorconstraints} constraints have particular meanings for the two dimensions: a vertical read direction and the horizontal direction for varation in time. This fact forms a special requirement for the layouting of operators.
Whereas the operator's positioning and resizing along the logical direction is only dependend on the sizes of sub operators, the arrangement within the time axis turns out to be more difficult. Here sub operators can't just be aligned centered as it is the case vertically. To face the determination of horizontal positioning, every operator gets time reference lines as depicted in figure~\ref{fig:referencelines}. Logical operators, i.e. IfThenOperator, AndOperator, OrOperator, NotOperator and StateOperator, have one reference line, time relevant operators, i.e. NextOperator, FutureOperator, AlwaysOperator and UntilOperator, have two such reference lines.

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{referencelines} 
  \caption{Time reference lines and their adjustment.}
  \label{fig:referencelines}
\end{figure}

%Über den eingebauten Layoutmanager-Mechanismus kann die Anordnung der Operatoren zueinander festgelegt werden. Wie aus section xy bekannt, gibt es eine logische wie auch eine zeitliche Achse bei constraints. Demnach muss die verticale wie auch horizontale positionierung der operatoren bestimmt werden. jeder operator besitzt eine logische Mittellinie, die wird zur mittellinie des übergeordneten Buckets ausgerichtet.

%DRAG&DROP

The inbuilt mouse support of swing was used for identifying operators laying under the current mouse position. On the one hand this is important for mouse sensitive highlighting of particular constraint parts, on the other hand this way it can be determined where operators have to be added after drag and drop. So far the predefined swing drag and drop has not been used but a simple drag and drop handler has been implemented instead. Once a JComponent is registered to the class DndHandler, it automatically manages any drag and drop actions with JComponents which implement either the Draggable or DropTarget interface. The former is implemented by all operators, the latter by buckets and the dashboard.
%Figure~\ref{fig:dndclassstructure} gives a short class overview.
Furthermore there is a class DraggableProvider which displays a kind of button in the tool bar and lets the user create new Draggables; in our case new operators get instantiated.

%das vorgefertigte Mouse-Handling wurde verwendet, um die jeweilige operator oder bucket componente an der aktuellen mausposition zu bestimmen. Die ist zum einen wichtig bei darüberfahrender maus um Teile des constraints hervorzuheben, zum anderen wird damit beim draggen auch bestimmt, welcher operator verschoben werden soll oder an welche stelle einer abgelegt werden soll.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{dndclassstructure} 
  \caption{Class definition of drag and drop functionality.}
  \label{fig:dndclassstructure}
\end{figure}

%operator werden im dashboard, einem weiteren container, zusammengeklickt und angezeigt. Das Dashboard hat dabei eine ähnliche Funktionsweise wie ein Bucket. die operatoren können dabei vom operatorprovider bezogen werden: Implementierungsdetails? trash...

\section{Automated constraint generation}
\label{sec:prototype:automatedconstraintgeneration}

@@

\section{Coalescence}
\label{sec:coalescence}

Both features the visual language presented in section~\ref{sec:operatorconstraints} and the automated constraint generation in chapter~\ref{chap:automatedconstraintgeneration} have to be assembled in one tool, the LTLCreator. It shall provide a dashboard with the possibility to manage several constraints, such as a tab panel, and displays for indicating validation results. Besides a button is needed for triggering constraint generation.

TODO: beschreiben dass vl im tab eingebunden wurde, die constraintgenerierung unter den zauberstab.

Figure~\ref{fig:editor} shows a snapshot of the editor's environment: All functionality needed for constraint editing is provided in the tool bar on the right. It contains draggable elements for creating all operator and proposition types as well as a trashcan for deleting. Constraints can be composed in the dashboard in the center of the editor by drag\&drop which holds and displays the operator constraints. The tab functionality on the left allows multiple constraints to be managed. Each tab shows a small thumbnail of the constraint and a symbol indicating its validity; a warning shield for ``incomplete'', a green shield for ``valid'', a red cross for ``invalid'' or an animated ring for ``validation in progress.''. The magic wand button within the tab pane brings the constraint generation functionality into the editor and triggers the constraint finding process.

\subsection{Model checker}

@@

a CTL model checker has been written in java, but...

Furthermore a particular model checker can be specified for constraint validation. As a default, the symbolic model checker NuSMV~\cite{springerlink:10.1007/s100090050046,NuSMV2} is used.

wie genau wird das getriggert (rekursiver listener an operatoren)

ConstraintValidator
ValidatorThread

(im hintergrund, mit warteschlange. dabei sollen aber wenn ein contsraint frequently geändert wird, alte validationsaufträge des gleichen constraints aus der warteschlange entfernt werden.
Callback mit result.

%\subsection{Integratability}
\subsection{Ease of integration}

LTLCreator should be available for other state machine based programming environments for integration what is ensured by its implementation as a java swing component and a clear API enabling the tool for a wide range of use.

Figure~\ref{fig:coalescence} points out how interaction between LTLCreator and programming environments where it is integrated in can be.
LTLCreator consists of a dashboard where constraints can be edited visually and an graphical output of validation results. The constraint generator as well as the validator are non UI parts of LTLCreator.
The programming environment provides the state machine model and notifies the LTLCreator component about all changes made to the program. This will automatically revalidate all existing constraints and always provide the up-to-date model for constraint generation.
For load and save functionality, all existing constraints can be received and new ones can be added to the LTLCreator component programmatically.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.7\linewidth]{coalescence}
  \caption{Interaction between LTLCreator and programming environment.}
  \label{fig:coalescence}
\end{figure}

\section{Integration into Robostudio}
\label{sec:integrationintorobostudio}

kurz erläutern, warum das gemaacht werden soll

Robostudio uses NetBeans as a platform, and all tools and features are organized in windows. Thus also LTLCreator has to be put into a window and added to the Robostudio platform. Since the defining of safety constraints is an important part not only after but also during the process of developing service robot behavior, the LTLCreator window as a vital tool shall be placed in the center of the platform where currently also the UI Component Layout Editor Window is. As already proposed in the previous section, all changes to the state machine program made within robostudio must of course cause the LTLCreator window to revalidate all existing constraints.
